OpenMPIR

### 2017-09-06 paper.tex
Addressing Billy's comments

\documentclass[sigconf]{acmart}
\newcommand{\wmnote}[1]{{\scriptsize \color{red} [[ Billy: #1]]}}

@@

Internal representations (IRs) are the \wmnote{the phrase compiler objects seems a bit odd to me}compiler objects on which almost all
analyses and optimizations are performed. Historically, a major drawback for
parallel codes is that IRs don't have the ability to represent, and therefore
reason about, parallelism. This prevents the application of common
optimizations to concurrent code. The recent work of Schardl et al.
\cite{shardl2017} has shown that real world compiler IRs can be extended to
represent fork-join parallelism. By extending the LLVM instruction set with
three instructions, they were able to capture all of the semantics to implement
the semantics of Cilk, an extension to C. They suggested that a similar
approach could be taken with OpenMP tasks.

@@

Internal representations are a crucial part of modern compilers. By representing
code in a generic, hardware-agnostic format, they allow both multiple frontends for
different source languages, and multiple backends for different hardware, to all
take advantage of a series of compiler optimizations and analyses.

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps LLVM has gained traction in the academic community for both its simplicity and relatively painless modification}LLVM is arguably the most successful IR to date \cite{llvm}. By implementing a simple
single static assignment (SSA) representation, it enables powerful optimizations 
that are effective for a wide range of programming languages and hardware
\cite{llvm imps}.

Historically, LLVM has had a sequential semantics: there is no notion of
concurrent or parallel processes. This resulted in existing compilers treating 
parallelism constructs as thin wrappers for calls into runtime systems. These
calls are essentially impossible for the compiler to reason about, due to the
complexity of runtimes being called.

This unfortunate situation was recently remedied by Schardl et al.
\cite{tapir}. By adding three instructions to LLVM, Tapir enables analyses and
optimizations of parallel code. A full description

@@

In this section we discuss a significant challenge in HPC: fragmentation of
parallel programming models. There are many, many, many ways to write
parallel programs \cite{....}. From libraries, to language extensions, to
standalone languages, to combinations of the above, it can be hard to choose
the right one. This means that when writing tooling, optimizations, or
analyses, one has to choose which model to target. One common difficulty among
all of these is that reasoning about parallel programs is hard \cite{}, and
compilers can help \cite{}. Therefore any duplicated work is extremely costly.
This is exactly analogous the problem LLVM helps to solve. Many optimizations
and analyses can be applied across a wide range of programming languages and 
hardware, but until there was an effective IR, those were often duplicated, 
resulting in a lot of lost opportunity.

Addressing these concerns this is a major motivation for our \wmnote{use of?}turning to Tapir.
\wmnote{this sentence feels very loose/run-on} By standardising a powerful but simple internal representation and building
frontends and backends for it, one both makes it easier to build optimizations
and analyses for a given programming model, and avoids duplicating work
across different implementations. We'll return to this issue later in the paper
when discussing potential limitations and challenges for Tapir for other
hardware and

@@

Specifically, Tapir introduces the \texttt{detach}, \texttt{reattach}, and
\texttt{sync} instructions. \wmnote{How in detail should I go here? I can talk
about these instructions on a broad level as a target for parallel tasks, or
more concretely define what they do. However, we don't seem discuss CFG's
before this so I'm not sure what level of introduction you're thinking of.}

@@

Tapir is implemented as an extension to the LLVM instruction set. Clang is a
C family compiler that has support for OpenMP extensions \cite{clang} and targets
LLVM. The existing Clang OpenMP implementation mapps OpenMP constructs directly
to OpenMP runtime library calls, by wrapping a C statement in a
\texttt{CapturedStmt}.  This replaces a C statement with a function call into
the OpenMP runtime library. along with a machine generated function who's body
contains that statement.

@@

While we did implement the codegen for the \texttt{task\_spawn} and
\texttt{task\_wait} constructs, it's worth noting here that even for these
pragmas, the implementation is incomplete. Currently, any clauses modifying the 
behaviour of the pragmas is ignored. Probably the most common semantics this
will change of variable semantics, e.g. \texttt{shared} vs. \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this issue would increase
the performance of this work marginally, due to reducing the number of memory
copies for variables declared \texttt{private} in OpenMP code.

The surprising fact that program behaviour wasn't changed significantly is
worth \wmnote{discussion?}visiting here. We surmise that it is likely due to the variable semantics
borrowed from the Cilk codegen and their relation to the OpenMP specification
of semantics.  In particular, \emph{OpenMP doesn't specify any memory model for
shared variables}. This results in the Cilk style of only writing to a shared
variable at the end of a parallel section as a perfectly valid implementation
of OpenMP's specification. In contrast, existing OpenMP implementations often
write to shared variables on every write. This is a potential for performance
discrepancy due to differing implementation-defined behaviour, and we will 
return to it when discussing performance on the Barcelona OpenMP task suite.
There is of course the fact that technically, by copying the value of variables
back to the surrounding context even for \texttt{private} variables, one isn't
following the specification. We surmise that in practice this hasn't had an
effect due to the fact that \texttt{private} variables are generally used for
performance, rather than their semantic properties.

A very important property to discuss of the current implementation is the
backend. This work has \emph{only been on the frontend}, which means we use
the only existing backend for Tapir. The existing backend is to generate 
calls into the Cilk runtime, \texttt{libcilkrts}. This obviously has many 
implications for adhering to the OpenMP specification. For example, environment
variables such as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by
\texttt{CILK\_NWORKERS}. We discuss some of these issues below, and return to 
address others in the future work section.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir as described above. Generally, a program has to contain other
OpenMP pragmas in order to run parallel code. A standard pattern for building
task-parallel OpenMP programs is to start a \texttt{parallel} pragma to spin up
the necessary hardware threads, then have a \texttt{single} pragma to have only
one thread continue on the specified statement. Then the other threads will get
work from spawned tasks, instead of implicitly executing the same code in a
data parallel style. For the purpose of this paper, we took the shortcut of
replacing these pragmas with no-ops. This worked well for all examples except
for one, in which after the \texttt{parallel} and \texttt{single} pragmas the
top level task was called with an unnecessary \texttt{task} pragma. Because the
wait was implicit and unhandled by our implementation, the simple fix was to
remove the \texttt{task} pragma. This was the only change required to the 
source code. Our temporary no-op shortcut, of course, doesn't follow the OpenMP
specification, and should be addressed by future work.

All code used for the implementation is available at \\
\texttt{https://github.com/lanl/openmpir}. The version used for this paper is
tagged as LLVM17.

@@

>>>>

\documentclass[sigconf]{acmart}
\newcommand{\wmnote}[1]{{\scriptsize \color{red} [[ Billy: #1]]}}
\newcommand{\gsnote}[1]{{\scriptsize \color{blue} [[ George: #1]]}}

@@

Compilers use internal representations (IRs) for implementing analyses and
optimizations. Historically, a major drawback for parallel programs is that IRs
don't have the ability to represent, and therefore reason about, parallelism.
This prevents the application of common optimizations to concurrent code. The
recent work of Schardl et al.  \cite{shardl2017} has shown that real world
compiler IRs can be extended to represent fork-join parallelism. By extending
the LLVM instruction set with three instructions, they were able to capture all
of the semantics to implement the semantics of Cilk, an extension to C. They
suggested that a similar approach could be taken with OpenMP tasks.

@@

\textbf{Internal representations (IRs)} are a crucial part of modern compilers.
By representing code in a generic, hardware-agnostic format, they allow both
multiple frontends for different source languages, and multiple backends for
different hardware, to all take advantage of a series of compiler optimizations
and analyses.

\wmnote{This feels like we're trying to convince someone about LLVM: perhaps
LLVM has gained traction in the academic community for both its simplicity and
relatively painless modification} 
\gsnote{I wanted to give a brief nod to the importance of LLVM, given the venue
(LLVM in HPC). Is this any more satisfactory?} 
Among IRs, LLVM is arguably the most ubiquitous \cite{llvm}. Winner of the
2012 ACM Software System Award, it is in use by frontends ranging from Haskell
to C++, and backends ranging from x86 to PTX \cite{}. Implementing a simple,
extensible \textbf{single static assignment (SSA)} IR, LLVM enables powerful
optimizations that are effective for a wide range of programming languages and
hardware \cite{llvm}.

Historically, LLVM has had a sequential semantics: there is no explicit notion
of concurrent or parallel processes. This has resulted in existing compilers
treating parallelism constructs as thin wrappers for calls into runtime
systems. These calls are infeasible for the compiler to reason about, due to
the complexity and flexibility of runtimes being called. This situation was
recently remedied by Schardl et al.  \cite{tapir}. By adding three instructions
to LLVM, Tapir enables analyses and optimizations of parallel code. We describe
Tapir in detail in Section~\ref{Sec:Tapir}. For a more complete description see
\cite{tapir}.

Arguably the most ubiquitous parallel programming model for on-node parallelism
in HPC, OpenMP has grown from being a way to implement parallel for loops to an
enormous and complex specification for parallelism. This paper focuses on a 
relatively recent addition to the OpenMP specification: the \texttt{task}
and \texttt{taskwait} constructs. A detailed description of these constructs
is presented in Section~\ref{Sec:OpenMP}. For a more complete description see
\cite{spec}.

@@

In this section we discuss a significant challenge in HPC: fragmentation of
parallel programming models. There are many, many, many ways to write
parallel programs \cite{....}. From libraries, to language extensions, to
standalone languages, to combinations of the above, fragmentation of parallel
programming models is a problem that is only getting worse. This means that
when writing tooling, optimizations, or analyses, one has to choose which model
to target. One common difficulty among all of these is that reasoning about
parallel programs is hard \cite{}, and compilers can help \cite{}. Therefore
any duplicated work is extremely costly.  This is exactly analogous the problem
LLVM helps to solve.  Many optimizations and analyses can be applied across a
wide range of programming languages and hardware, but until there was an
effective IR, those were often duplicated, resulting in a lot of lost
opportunity.

Addressing these concerns this is a major motivation for our use of Tapir.
By standardising an IR, in addition to easing the development of optimizations,
one avoids duplicating work across different implementations. We'll return to
this issue later in Section~\ref{Sec:Future} when discussing future work.

@@

Specifically, Tapir introduces the \texttt{detach}, \texttt{reattach}, and
\texttt{sync} instructions. \wmnote{How in detail should I go here? I can talk
about these instructions on a broad level as a target for parallel tasks, or
more concretely define what they do. However, we don't seem discuss CFG's
before this so I'm not sure what level of introduction you're thinking of.}
\gsnote{I was thinking roughly a page should be enough. Maybe a summary of what
Tapir is followed by an example of an optimization it enables. I like both the
CSE and tail-call optimization examples from your PPoPP paper. Feel free to 
define CFG and any other terms you need to.}

@@

Tapir is implemented as an extension to the LLVM instruction set. Clang is a
C family compiler that has support for OpenMP extensions \cite{clang} and targets
LLVM. The existing Clang OpenMP implementation maps OpenMP constructs directly
to OpenMP runtime library calls, by wrapping a C statement in a
\texttt{CapturedStmt}. This replaces a C statement with a function call into
the OpenMP runtime library, along with a machine generated function who's body
contains that statement.

@@

While we did implement the codegen for the \texttt{task\_spawn} and
\texttt{task\_wait} constructs, it's worth noting here that even for these
pragmas, the implementation is incomplete. Currently, any clauses modifying the 
behaviour of the pragmas is ignored. Probably the most common semantics this
will change of variable semantics, e.g. \texttt{shared} vs. \texttt{private}.
Surprisingly, this had little effect on the correctness of the entire Barcelona
OpenMP task suite. Indeed, it is likely that fixing this issue would increase
the performance of this work marginally, due to reducing the number of memory
copies for variables declared \texttt{private} in OpenMP code. This property
that behaviour wasn't changed significantly is worth revisiting, and we do so
in Section~\ref{Sec:Discussion}.

A very important property of the current implementation to discuss is the
backend. This work has \emph{only been on the frontend}, which means we use
the only existing backend for Tapir. The existing backend generates calls into
the Cilk runtime, \texttt{libcilkrts}. This obviously has many implications for
adhering to the OpenMP specification. For example, environment variables such
as \texttt{OMP\_NUM\_THREADS} are ignored, replaced by \texttt{CILK\_NWORKERS}.
We discuss some of these issues below, and return to address others in the
future work section.

There are other issues that had to be overcome to run full OpenMP programs
using Tapir as described above. Generally, in addition the the \texttt{task} 
and \texttt{taskwait} pragmas, a program has to contain initialization OpenMP
pragmas in order to run parallel code. A standard pattern for building
task-parallel OpenMP programs is to insert a \texttt{parallel} pragma to start
the necessary hardware threads, followed immediately by a \texttt{single}
pragma to have only one thread continue on the specified statement. Then the
other threads will get work from spawned tasks, instead of implicitly executing
the same code in a data parallel style. For the purpose of this paper, we took
the shortcut of replacing these pragmas with no-ops. This worked well for all
examples except for one, in which after the \texttt{parallel} and
\texttt{single} pragmas the top level task was called with an unnecessary
\texttt{task} pragma.  Because the wait was implicit and unhandled by our
implementation, the simple fix was to remove the \texttt{task} pragma. This was
the only change required to the source code. Our temporary no-op shortcut, of
course, doesn't follow the OpenMP specification, and should be addressed by
future work.

All code used for the implementation is available at \\
\texttt{https://github.com/lanl/openmpir-clang}. The version used for this
paper is tagged as LLVM17.

@@

The surprising fact that program behaviour wasn't changed significantly is
worth discussing here. We surmise that it is likely due to the variable
semantics borrowed from the Cilk codegen and their relation to the OpenMP
specification of semantics. In particular, \emph{OpenMP doesn't specify any
memory model for shared variables}. This results in the Cilk style of only
writing to a shared variable at the end of a parallel section as a perfectly
valid implementation of OpenMP's specification. In contrast, existing OpenMP
implementations often write to shared variables on every write. This is a
potential for performance discrepancy due to differing implementation-defined
behaviour, and we will return to it when discussing performance on the
Barcelona OpenMP task suite. There is of course the fact that technically, by
copying the value of variables back to the surrounding context even for
\texttt{private} variables, one isn't following the specification. We surmise
that in practice this hasn't had an effect due to the fact that
\texttt{private} variables are generally used for performance, rather than
their semantic properties.
